\documentclass[journal]{IEEEtran}

% ---------------------------------------------------------------------
% Packages
% ---------------------------------------------------------------------
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{algorithmic}
\usepackage{array}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\usepackage{textcomp}
\usepackage{stfloats}
\usepackage{url}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage[T1]{fontenc}
\usepackage{tikz}
\usetikzlibrary{trees,positioning,shapes.geometric,arrows.meta,fit,backgrounds}
\usepackage[edges]{forest}
\usepackage{cite}
\usepackage{balance}

\hyphenation{op-tical net-works semi-conduc-tor IEEE-Xplore}

% ---------------------------------------------------------------------
% Colors used by the taxonomy figure
% ---------------------------------------------------------------------
\definecolor{axisA}{RGB}{31,119,180}
\definecolor{axisB}{RGB}{44,160,44}
\definecolor{axisC}{RGB}{148,103,189}
\definecolor{rootgray}{RGB}{80,80,80}

\begin{document}

\title{A Survey on Reinforcement Learning for \\ LLM Training}

\author{Gustavo~Pantuza,
        Daniel~Neves,
        Nat\'alia~Sales,
        Caio~Campos,
        Edwaldo~Rodrigues
        
\thanks{The authors are with the Department of Computer Science,
Universidade Federal de Minas Gerais (UFMG), Belo Horizonte, Brazil
(e-mail: \{pantuza, daniel.neves, natalia.sales, caio.campos, edwaldo.rodrigues\}@dcc.ufmg.br).}
\thanks{Manuscript prepared June 2026.}}

\markboth{IEEE Communications Surveys \& Tutorials,~Vol.~XX, No.~X, 2026}%
{Campos \MakeLowercase{\textit{et al.}}: A Survey on Reinforcement Learning for LLM Training}

\maketitle

\begin{abstract}
Reinforcement learning (RL) has become the central mechanism for
transforming large language models (LLMs) from passive next-token
predictors into capable problem solvers and autonomous decision-making
agents. From the alignment of instruction-following assistants with
human preferences to the emergence of long-horizon reasoning in
models trained with verifiable rewards, RL now spans the entire
post-training stage of the modern LLM lifecycle. However, the field
has grown rapidly and unevenly, with overlapping terminology, diverse
algorithms, and heterogeneous reward designs that make systematic
comparison difficult. This survey provides a comprehensive, tutorial
review of RL for LLM training. We first formalize the paradigm shift
from single-step \emph{preference-based reinforcement fine-tuning}
(PBRFT) to temporally extended \emph{agentic RL} through a unified
Markov decision process lens, and recall the algorithmic spectrum
spanning REINFORCE, PPO, DPO, and GRPO. We then introduce a compact
yet expressive three-axis taxonomy that organizes the literature by
\emph{(i)} the reward signal that drives learning, \emph{(ii)} the
optimization algorithm that updates the policy, and \emph{(iii)} the
training scope that determines whether a single response or a full
multi-step trajectory is optimized. Using this taxonomy, we map
representative works, contrast their trade-offs, and consolidate the
landscape of open challenges, including reward hacking, credit
assignment, sample efficiency, and scalable oversight. We conclude by
outlining research directions toward reliable and general-purpose
agentic intelligence.
\end{abstract}

\begin{IEEEkeywords}
Large language models, reinforcement learning, RLHF, RLAIF,
direct preference optimization, GRPO, reward modeling, LLM agents,
alignment, taxonomy.
\end{IEEEkeywords}

\input{01-introduction}
\input{02-background}
\input{03-taxonomy}
\input{04-reward-signals}
\input{05-optimization}
\input{06-agentic-rl}
\input{07-challenges}
\input{08-conclusion}

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
